Estimating the number of clusters using Cross Validation

### 2015-06-21 cvclust.tex

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sbA}{\mathcal{\bar A}}
\newcommand{\sC}{\mathcal{C}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sN}{\mathcal{N}}
\newcommand{\sX}{\mathcal{X}}

@@

\subsection*{Gabriel CV algorithm}

@@

\begin{theorem}[$k$-Means Consistency]\label{lem:kmeans-consistency}
Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes $W(\cdot,
P_n)$ over $\sA_k$.  Then for all $n$ there exists an $\bar A_n \in \sbA_k$
such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$, and $W(A_n, P_n) \to
m_k(P)\text{ a.s.}$ \end{theorem}

@@

\begin{lemma}
\label{lem:pollard-compact}
Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If $W(B_n,
P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then there exists a finite
$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in the
class of sets
\(
  \sE_k = \{ A \subseteq B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}

\begin{proof}
Pollard proves this result, in Section~3 of his paper,  assuming that there is
a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of $j = 1,
\dotsc, k$.  His proof, which is very similar to the proof of
Lemma~\ref{lem:kmeans-existence}, does not use this assumption directly, only
its implication that $m_{k}(P) < m_{k-1}(P)$.  Since we have established his
result in Lemma~\ref{lem:kmeans-existence}, the remainder of his proof applies
without further modification.
\end{proof}

@@

With Lemmas~\ref{lem:kmeans-continuous}--\ref{lem:kmeans-uniform-slln}, we can
proceed to the proof of Theorem~\ref{lem:kmeans-consistency}.

\begin{proof}[Proof of Theorem~\ref{lem:kmeans-consistency}]
The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
$\sbA_k$.

@@

\begin{lemma}
Set $\mu = \int x P(dx)$.  If $\int \|x\|^2 P(dx) < \infty$, then there exists
an $r > 0$ such that for every $A \in \sbA_k$, there is at most one center $a
\in A$ with $\|a - \mu\| \leq r$.
\end{lemma}
\begin{proof}
If the support of $P$ has fewer than $k$ points, then the result is trivial.
Otherwise, Lemma~\ref{lem:kmeans-existence} implies that $m_k(P) <
m_{k-1}(P)$.  Define
\[
  d =
    \inf_{A \in \sA_k}
      \min \{ \|a - a'\| : a,a' \in A, a \neq a' \}.
\]
If $d = 0$, then there exists a sequence $A_n$ in $\sbA_k$ with
\[
  \min_{\substack{a,a' \in A_n \\ a \neq a'}} \|a - a'\| \to 0.
\]
Since $\sbA_k$ is compact, there exists a subsequence $n(m)$ and a set $A_0
\in \sbA_k$ with $A_{n(m)} \to A_0$.  Take $B_n \in \sA_{k-1}$ to be the set
formed by removing one of the centers $a'$ minimizing $\min_{\substack{a \in
A_n \\ a \neq a'}} \|a - a'\|$.  By construction, $H(A_n, B_n) \to 0$, so that
$B_{n(m)} \to A_0$.  Since the map $A \mapsto W(A, P)$ is continuous on
$\sA_k$, we must have $W(B_{n(m)}, P) \to W(A_0, P) = m_k(P)$.  However, this
is impossible since $W(B_n,P) \geq m_{k-1}(P) > m_k(P)$.  It must be the case,
then, that $d > 0$.

@@

Observations $\{ (X_i, Y_i) \}_{i=1}^{n+N}$ are independent and identically
distributed.  The distribution for each $(X_i, Y_i)$ pair is such that $X_i$
and $Y_i$ have marginal distribution $P^X$ and $P^Y$, and the vectors $X_i$
and $Y_i$ are independent.

>>>>

% operators
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}

% convergence
\newcommand{\toas}{\overset{\mathit{a.s.}}{\to}}

% sets
\newcommand{\R}{\mathbb{R}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sbA}{\mathcal{\bar A}}
\newcommand{\sC}{\mathcal{C}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sN}{\mathcal{N}}
\newcommand{\sX}{\mathcal{X}}

% scalars
\newcommand{\bpi}{\bar \pi}

% vectors
\newcommand{\muX}{\mu^{X}}
\newcommand{\muY}{\mu^{Y}}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}

% matrices
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\SigmaY}{\Sigma^Y}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}

% class labels
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}

@@

\subsection{Gabriel CV algorithm}
\label{sec:gabriel-cv-algorithm}

@@

\begin{theorem}[$k$-Means Consistency]\label{thm:kmeans-consistency}
Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes $W(\cdot,
P_n)$ over $\sA_k$.  Then for all $n$ there exists an $\bar A_n \in \sbA_k$
such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$, and $W(A_n, P_n) \to
m_k(P)\text{ a.s.}$ \end{theorem}

@@

\begin{lemma}
\label{lem:pollard-compact}
Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If $W(B_n,
P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then there exists a finite
$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in the
class of sets
\(
  \sE_k = \{ A \subseteq B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}
\begin{proof}
Pollard proves this result, in Section~3 of his paper,  assuming that there is
a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of $j = 1,
\dotsc, k$.  His proof, which is very similar to the proof of
Lemma~\ref{lem:kmeans-existence}, does not use this assumption directly, only
its implication that $m_{k}(P) < m_{k-1}(P)$.  Since we have established his
result in Lemma~\ref{lem:kmeans-existence}, the remainder of his proof applies
without further modification.
\end{proof}

@@

With Lemmas~\ref{lem:kmeans-continuous}--\ref{lem:kmeans-uniform-slln}, we can
proceed to the proof of Theorem~\ref{thm:kmeans-consistency}.

\begin{proof}[Proof of Theorem~\ref{thm:kmeans-consistency}]
The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains
$\sbA_k$.

@@

\begin{lemma}\label{lem:kmeans-separation}
Set $\mu = \int x P(dx)$.  If $\int \|x\|^2 P(dx) < \infty$, then there exists
an $r > 0$ such that for every $A \in \sbA_k$, there is at most one center $a
\in A$ with $\|a - \mu\| \leq r$.
\end{lemma}
\begin{proof}
If the support of $P$ has fewer than $k$ points, then the result is trivial.
Otherwise, Lemma~\ref{lem:kmeans-existence} implies that $m_k(P) <
m_{k-1}(P)$.  Define
\[
  d =
    \inf_{A \in \sA_k}
      \min \{ \|a - a'\| : a,a' \in A, a \neq a' \}.
\]
If $d = 0$, then there exists a sequence $A_n$ in $\sbA_k$ with
\[
  \min_{\substack{a,a' \in A_n \\ a \neq a'}} \|a - a'\| \to 0.
\]
Since $\sbA_k$ is compact, there exists a subsequence $n(m)$ and a set $A_0
\in \sbA_k$ with $A_{n(m)} \to A_0$.  Take $B_n \in \sA_{k-1}$ to be the set
formed by removing one of the centers $a'$ minimizing $\min_{\substack{a \in
A_n \\ a \neq a'}} \|a - a'\|$.  By construction, $H(A_n, B_n) \to 0$, so that
$B_{n(m)} \to A_0$.  Since the map $A \mapsto W(A, P)$ is continuous on
$\sA_k$, we must have $W(B_{n(m)}, P) \to W(A_0, P) = m_k(P)$.  However, this
is impossible since $W(B_n,P) \geq m_{k-1}(P) > m_k(P)$.  It must be the case,
then, that $d > 0$.

@@

\begin{theorem}
For each $n$, let $N(n)$ be a nonnegative integer with $N(n) \to \infty$ as $n
\to \infty$, and let $\{ (X_{n,i}, Y_{n,i}) \}_{i=1}^{n+N(n)}$ be independent
identically distributed random variables.  For any fixed $k$, let $\CV(k)$
be the cross-validation error from a single Gabriel CV fold, computed as
described in Sec.~\ref{sec:gabriel-cv-algorithm}.  Suppose that
\begin{enumerate}[label=(\roman*)]
\item the distribution for each $(X, Y)$ pair is such that $X$ and $Y$ have
  marginal distributions $P^X$ and $P^Y$, and the vectors $X$ and $Y$ are
  independent of each other;

\item the distribution $P^Y$ satisfies $\int \|y\|^2 P^Y(dy) < \infty$;

\item the support of $P^Y$ contains at least two points.
\end{enumerate}
Then for any fixed $k > 1$, almost surely eventually  $\CV(k) > \CV(1)$.
\end{theorem}
\begin{proof}
Set $\muY = \E(Y)$ and $\SigmaY = \cov(Y)$.  The strong law of large
numbers implies that
\(
  \CV(1) \toas \int \|y - \mu^Y\|^2 P(dy) = \tr(\SigmaY).
\)
If $k > 1$, then Theorem~\ref{thm:kmeans-consistency} and
Lemma~\ref{lem:kmeans-separation} imply that there exists an $r > 0$ such that
almost surely eventually at most one cluster center in
$\{ \bmuY_1, \dotsc, \bmuY_k \}$ is within distance $r$ of $\muY$.  Without
loss of generality suppose that $\bmuY_1$ is the closest cluster center to
$\muY$.
\end{proof}
